A Review of Neutron Imaging

### 2020-10-09 neutron-imaging/neutron.tex
WIP 09Oct2020

\numberwithin{equation}{section}
\linespread{1.3}
\title{A Review of Neutron Imaging}
\author{To be filled}
\date{27-Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s1}
Nobody suspected the presence of a neutral particle in an atomic nucleus until
the 1930s. When Heisenberg developed matrix mechanics in 1925 and 
Schr\"{o}dinger wrote his celebrated equation, the nucleus was thought to be
made up of positively charged protons and a few negatively charged `internal
electron'. The latter were supposed to be in present in sufficient numbers to
account for the mass and charge of the nucleus. However, this idea became
increasingly untenable with the development of quantum statistical mechanics
and an understanding of Fermions and Bosons. When Oscar Klein 
\cite{klein1929reflexion} applied the newly discovered Dirac equation to the
`internal electrons' in the nucleus he found that they should find it 
extremely easy to escape the potential barrier of the proton's electric field.
These difficulties were finally resolved when Chadwick reported 
\cite{chadwick1932existence}, the existence of an electrically neutral particle, called the neutron, in the atomic nucleus.

\section{Physical properties of a neutron}\label{s2}
The mass of a free neutron is $1.674927471 \times 10^{-27}$ kg
\cite{taylor2007nist}. Its mean square radius is $0.8 \times 10^{-15}$ m
\cite{bogdan2015particles}. It has no electric charge. However, it has
an intrinsic spin of $1/2$ (in units of $\hslash$)
\cite{basdevant2005fundamentals}. As a result, although it is unaffected by 
an electric field, it does respond to a magnetic field. In fact, a neutron 
responds to all four fundamental forces in Nature. Outside of
the nucleus, the neutron is unstable. It decays into a proton, an electron and
an electron neutrino with a half life of about $10$ minutes and $10$ seconds
\cite{Nakamura_2010}. The energy of decay is $0.782$ MeV \cite{heyde2004basic}.
A small fraction, about $0.001$, of free neutrons also produce gamma radiation 
while decaying. An even smaller fraction, about $4$ per million, of neutrons
produce electrons with an energy lesser than $13.6$ eV. As a result, they
stay stay bound to the proton forming a hydrogen atom. The fact that neutrons
decay cannot be ignored while using them for imaging. This feature of the
neutrons distingushes them from the other probes used for imaging like optical
photons, X-rays, gamma rays, infra-red radiation, sound waves and electrons. 
The latter probes are not known to decay.

@@

\section{Interaction of neutrons with matter}\label{s4}
Neutrons being electrically neutral can travel significant distances in
condensed matter. The probability $\mu$ that a neutron interacts with matter
in unit distance is called its linear attenuation coefficient. It depends on
the energy of the neutron and the properties of the condensed matter through
which it is traveling. If $N(x)$ is the number of neutron that have traversed
a distance $x$ in the condensed matter without any interaction then the number
of neutrons $dN$ that will interact in the distance interval $x$ and $x + dx$
is
\begin{equation}\label{s4e1}
dN = -\mu Ndx.
\end{equation}
The solution of this equation is 
\begin{equation}\label{s4e2}
N = N_0 e^{-\mu x},
\end{equation}
where $N_0 = N(x = 0)$ is the number of neutrons incident on a beam of the
material. Neutrons interact only with atomic nuclei. In an elastic collision
they transfer a portion of their kinetic energy to the nuclei. A series of
elastic collisions leads to slowing of neutrons, also called their moderation.
Eventually, the kinetic energy of the neutron drops down to the level of the
thermal energy of the surroundings and it gets absorbed by a nucleus. Neutrons
can also undergo inelastic collisions with nuclei. In this case, the neutron
transfers a portion of its energy to the incident nucleus, raising it to an 
excited state. Even in this case, neutron meets the same fate of losing energy
up to the level of thermal energy and getting absorbed. Materials rich in 
protons, that is H${}^+$ ions, show an overwhelmingly large proportion of
elastic collisions because the bare proton does not have excited states. The
de-energized neutron is absorbed leading the formation of a Deuterium atom
\begin{equation}\label{s4e3}
{}_1^1H + {}^1_0 n \rightarrow {}^2_1 H + \gamma.
\end{equation}
The energy of the resulting $\gamma$-radiation is $2.22$ MeV. The precise 
energy of this radiation is used to detect the presence of neutrons.
In
materials with other nuclei, a plot of the neutron collision cross sections 
with neutron energies show pronounced peaks and troughs at the nuclear 
transition levels. A notable aspect of elastic collisions is that a neutron of
mass $m_n$ can transfer its kinetic energy to a nucleus of mass $m$ to the
extent of
\begin{equation}\label{s4e4}
Q = \frac{4m_nmE_n}{(m + m_n)^2},
\end{equation}
where $E_n$ is the energy of the neutrons. In a material rich with protons,
$m = m_n$ and $Q = 1$. Light nuclei are thus far more effective in slowing
neutrons than heavier ones. Elastic collisions of neutrons with heavy nuclei
change the direction of motion of nuclei while absorbing only a small portion
of their kinetic energy.
\section{Detection of neutrons}\label{s5}
Neutrons are chargeless and their magnetic moment is too low to be used for
reliably detecting their presence. They also have a small electric dipole 
moment but it has not been measured in a laboratory. Free neutrons decay to
produce a proton and an electron but its half life of around $10$ minutes is
too long to permit the detection of neutrons through its decay products. That
leaves us with just one option, consider nuclear reactions triggered by the
neutrons and measure their results to estimate the number and energy of the
neutrons. Since the imaging applications involve mostly slow neutrons, we will
focus our attention on their detection.

Slow neutrons are detected by the nuclear reactions they trigger in ${}^{10}_5$
B, ${}^6_3$Li and ${}^3_2$He. The first two of these produce alpha particles
while the last one produces hydrogen atom. One can use non-relativistic energy
and momentum conservation laws to show that the energy of the alpha particle
in these reactions is \cite{turner2008atoms}
\begin{equation}\label{s5e1}
E = \frac{MQ}{m + M},
\end{equation}
where $m$ is the mass of the alpha particle, $M$ that of the daughter 
nucleus and $Q$ is the energy released in the reaction. The resulting alpha
particles are detected using proportional counters. Proportional counters
measure only the number of neutrons, not their energy spectrum. The reactions
used for detection are \cite{crane1991neutron}
\begin{equation}\label{s5e2}
{}^{10}_5\text{B}+{}^1_0\text{n} \rightarrow \begin{cases}
{}^7_3\text{Li}^\star + {}^4_2\text{He} + 2.31\text{ MeV} \\
{}^7_3\text{Li} + {}^4_2\text{He} + 2.79\text{ MeV}, 
\end{cases}
\end{equation}
where the $\star$ superscript in the first reaction indicates that the nucleus
is in an excited state. The other reactions are
\begin{eqnarray}
{}^6_3\text{Li}+{}^1_0\text{n}&\rightarrow&{}^3_1\text{H} + {}^4_2\text{He} +
4.78\text{ MeV} \label{s5e3} \\
{}^3_2\text{He}+{}^1_0\text{n}&\rightarrow&{}^3_1\text{H} + {}^1_1\text{H} +
0.765\text{ MeV} \label{s5e4} 
\end{eqnarray}
\section{Neutron dosimetry and shielding}\label{s6}
\section{Propagation of neutron beams}\label{s7}
\section{Neutron diffraction by crystals}\label{s8}
\section{2-d neutron imaging}\label{s9}
\section{Neutron tomography}\label{s10}
\section{Imaging of biologically important structures}\label{s11}
\section{Larmor precession based imaging}\label{s12}
\section{Polarization based imaging}\label{s13}
\section{Conclusion}\label{s14}
\bibliography{neutron}
\bibliographystyle{plain}
\end{document}

>>>>

\numberwithin{equation}{section}
\linespread{1.3}
\title{A Review of Neutron Imaging}
\author{To be filled}
\date{27-Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s1}
Nobody suspected the presence of a neutral particle in an atomic nucleus until
the 1930s. When Heisenberg developed matrix mechanics in 1925 and 
Schr\"{o}dinger wrote his celebrated equation, the nucleus was thought to be
made up of positively charged protons and a few negatively charged `internal
electrons'. The latter were supposed to be in present in sufficient numbers to
account for the mass and charge of the nucleus. However, this idea became
increasingly untenable with the development of quantum statistical mechanics
and an understanding of Fermions and Bosons. When Oscar Klein 
\cite{klein1929reflexion} applied the newly discovered Dirac equation to the
`internal electrons' in the nucleus he found that they should find it 
extremely easy to escape the potential barrier of the proton's electric field.
These difficulties were finally resolved when Chadwick reported 
\cite{chadwick1932existence}, the existence of an electrically neutral particle, called the neutron, in the atomic nucleus.

\section{Physical properties of a neutron}\label{s2}
The mass of a free neutron is $1.674927471 \times 10^{-27}$ kg
\cite{taylor2007nist}. At non-relativistic speed $v$, its de Broglie wavelength
is \cite{anderson2008neutron}
\begin{equation}\label{s2e1}
\lambda = \frac{395.6}{v} \text{nm}
\end{equation}
making it an ideal source to explore condensed matter at the atomic scale.

Its mean square radius is $0.8 \times 10^{-15}$ m
\cite{bogdan2015particles}. It has no electric charge. However, it has
an intrinsic spin of $1/2$ (in units of $\hslash$)
\cite{basdevant2005fundamentals}. As a result, although it is unaffected by 
an electric field, it does respond to a magnetic field. In fact, a neutron 
responds to all four fundamental forces in Nature. Outside of
the nucleus, the neutron is unstable. It decays into a proton, an electron and
an electron neutrino with a half life of about $10$ minutes and $10$ seconds
\cite{Nakamura_2010}. The energy of decay is $0.782$ MeV \cite{heyde2004basic}.
A small fraction, about $0.001$, of free neutrons also produce gamma radiation 
while decaying. An even smaller fraction, about $4$ per million, of neutrons
produce electrons with an energy lesser than $13.6$ eV. As a result, they
stay stay bound to the proton forming a hydrogen atom. The fact that neutrons
decay cannot be ignored while using them for imaging. This feature of the
neutrons distingushes them from the other probes used for imaging like optical
photons, X-rays, gamma rays, infra-red radiation, sound waves and electrons. 
The latter probes are not known to decay.

@@

\section{Interaction of neutrons with matter}\label{s4}
Neutrons being electrically neutral can travel significant distances in
condensed matter. The probability $\mu$ that a neutron interacts with matter
in unit distance is called its linear attenuation coefficient. It depends on
the energy of the neutron and the properties of the condensed matter through
which it is traveling. If $N(x)$ is the number of neutron that have traversed
a distance $x$ in the condensed matter without any interaction then the number
of neutrons $dN$ that will interact in the distance interval $x$ and $x + dx$
is
\begin{equation}\label{s4e1}
dN = -\mu Ndx.
\end{equation}
The solution of this equation is 
\begin{equation}\label{s4e2}
N = N_0 e^{-\mu x},
\end{equation}
where $N_0 = N(x = 0)$ is the number of neutrons incident on a beam of the
material. Neutrons interact only with atomic nuclei. In an elastic collision
they transfer a portion of their kinetic energy to the nuclei. A series of
elastic collisions leads to slowing of neutrons, also called their moderation.
Eventually, the kinetic energy of the neutron drops down to the level of the
thermal energy of the surroundings and it gets absorbed by a nucleus. Neutrons
can also undergo inelastic collisions with nuclei. In this case, the neutron
transfers a portion of its energy to the incident nucleus, raising it to an 
excited state. Even in this case, neutron meets the same fate of losing energy
up to the level of thermal energy and getting absorbed. Materials rich in 
protons, that is H${}^+$ ions, show an overwhelmingly large proportion of
elastic collisions because the bare proton does not have excited states. The
de-energized neutron is absorbed leading the formation of a Deuterium atom
\begin{equation}\label{s4e3}
{}_1^1H + {}^1_0 n \rightarrow {}^2_1 H + \gamma.
\end{equation}
The energy of the resulting $\gamma$-radiation is $2.22$ MeV. The precise 
energy of this radiation is used to detect the presence of neutrons.
In
materials with other nuclei, a plot of the neutron collision cross sections 
with neutron energies show pronounced peaks and troughs at the nuclear 
transition levels. A notable aspect of elastic collisions is that a neutron of
mass $m_n$ can transfer its kinetic energy to a nucleus of mass $m$ to the
extent of
\begin{equation}\label{s4e4}
Q = \frac{4m_nmE_n}{(m + m_n)^2},
\end{equation}
where $E_n$ is the energy of the neutrons. In a material rich with protons,
$m = m_n$ and $Q = 1$. Light nuclei are thus far more effective in slowing
neutrons than heavier ones. Elastic collisions of neutrons with heavy nuclei
change the direction of motion of nuclei while absorbing only a small portion
of their kinetic energy.

The interaction of neutrons with matter are used to develop neutron imaging
techniques. The elastic scattering gives information about the atomic 
structure. Neutron diffraction techniques have been used to study solids. On
the other hand, inelastic scattering reveals information about collective
excitations. Interactions of both kinds give rise to contrast that is used
for imaging.

\section{Detection of neutrons}\label{s5}
Neutrons are chargeless and their magnetic moment is too low to be used for
reliably detecting their presence. They also have a small electric dipole 
moment but it has not been measured in a laboratory. Free neutrons decay to
produce a proton and an electron but its half life of around $10$ minutes is
too long to permit the detection of neutrons through its decay products. That
leaves us with just one option, consider nuclear reactions triggered by the
neutrons and measure their results to estimate the number and energy of the
neutrons. Since the imaging applications involve mostly slow neutrons, we will
focus our attention on their detection.

Slow neutrons are detected by the nuclear reactions they trigger in ${}^{10}_5$
B, ${}^6_3$Li and ${}^3_2$He. The first two of these produce alpha particles
while the last one produces hydrogen atom. One can use non-relativistic energy
and momentum conservation laws to show that the energy of the alpha particle
in these reactions is \cite{turner2008atoms}
\begin{equation}\label{s5e1}
E = \frac{MQ}{m + M},
\end{equation}
where $m$ is the mass of the alpha particle, $M$ that of the daughter 
nucleus and $Q$ is the energy released in the reaction. The resulting alpha
particles are detected using proportional counters. Proportional counters
measure only the number of neutrons, not their energy spectrum. The reactions
used for detection are \cite{crane1991neutron}
\begin{equation}\label{s5e2}
{}^{10}_5\text{B}+{}^1_0\text{n} \rightarrow \begin{cases}
{}^7_3\text{Li}^\star + {}^4_2\text{He} + 2.31\text{ MeV} \\
{}^7_3\text{Li} + {}^4_2\text{He} + 2.79\text{ MeV}, 
\end{cases}
\end{equation}
where the $\star$ superscript in the first reaction indicates that the nucleus
is in an excited state. The other reactions are
\begin{eqnarray}
{}^6_3\text{Li}+{}^1_0\text{n}&\rightarrow&{}^3_1\text{H} + {}^4_2\text{He} +
4.78\text{ MeV} \label{s5e3} \\
{}^3_2\text{He}+{}^1_0\text{n}&\rightarrow&{}^3_1\text{H} + {}^1_1\text{H} +
0.765\text{ MeV} \label{s5e4} 
\end{eqnarray}
We mentioned previously that the collision between a proton and a neutron is
always elastic. Further, since their masses are almost equal, it is possible
for the neutron to transfer all its kinetic energy to the proton. The proton
is then detected by scintillators. Organic compounds are usually rich in 
protons and detectors based on this principle use them in crystalline, 
amorphous or even liquid forms.

Bubble detectors are inexpensive and portable neutron detecting devices. They
consist of a transparent polymer with a large number of tiny droplets of 
a superheated liquid. The passage of a charged particle through such droplets
results in a rapid phase change from liquid to vapor. The transparent droplets
become visible bubbles which are counted either manually or electronically.
\section{Neutron dosimetry and shielding}\label{s6}
Radiation Dosimetry is the measurement of the `dose' of radiation absorbed by
a biological sample. Dosimetry is important for the safety of people working
with radiation. Biological materials are rich in protons. When they are
exposed to neutrons, the elastic collision of the neutrons and the protons
leads to a transfer of a substantial amount of energy to the protons resulting
in breaking of chemical bonds. Neutrons damage biological samples more than
a comparable dose of electrons or $\gamma$ rays. The amount of energy absorbed
by $1$ kg of a material is called the `absorbed dose'. It is measured in grays,
where $1$ gray is equal to $1$ Jkg${}^{-1}$. It is measured and specified at 
every point of the exposed sample. Before getting into the details of neutron
dosimetry, we introduce a few physical concepts.

The restricted stopping power of a target is the the linear rate of energy of
energy lost,  $(-dE/dx)_\Delta$, due to collisions alone and in which the
energy transfer does not exceed $\Delta$. If $\mu$ is the probability per unit
distance of travel that a neutron collides with a proton and if $W(Q)dQ$ is the 
probability that a given collision will result in an energy loss between $Q$
and $Q + dQ$ then
\begin{equation}\label{s6e1}
-\left(\frac{dE}{dx}\right)_\Delta = \mu\int_{Q_{\text{min}}}^\Delta QW(Q)dQ,
\end{equation}
where $Q_{\text{min}}$ is the minimum energy needed for excitation of the 
target \cite{turner2008atoms}. The stopping power depends on both the 
incident radiation and the absorbing sample. Radiation with high restricted
stopping power is more dangerous than the one with low restricted stopping 
power. The product of the absorbed dose $D$ and a quality factor $Q$ (not 
related to the $Q$ in equation \eqref{s6e1} is called the `dose equivalent'
$H$ of the sample. The absorbed dose is also a point function of the sample.
That is, it can be specified at every point of the sample. The SI unit of 
absorbed dose is sievert.

Neutrons are best shielded by material rich in protons like Iron, water, 
paraffin, concrete and graphite, in decreasing order of attenuation length
\cite{turner2008atoms}.
\section{Propagation of neutron beams}\label{s7}
\section{Neutron diffraction by crystals}\label{s8}
\section{2-d neutron imaging}\label{s9}
\section{Neutron tomography}\label{s10}
\section{Imaging of biologically important structures}\label{s11}
\section{Larmor precession based imaging}\label{s12}
\section{Polarization based imaging}\label{s13}
\section{Conclusion}\label{s14}
\bibliography{neutron}
\bibliographystyle{plain}
\end{document}
